Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-05-16 TEBenchmarkPaper/article.tex
TEB: Close many issues

% Derived from the template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{csvsimple}
\usepackage{float}
\restylefloat{table}
\usepackage{hhline}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{arrows, calc, decorations.pathreplacing, positioning,
  shapes.geometric}
% etc.

@@

\begin{itemize}
\item A general methodology for benchmarking conjecture generation.
\item Resolving the issue of ``interestingness'' through the use of
  theorem-proving benchmarks as a ground-truth.
\item A specific instantiation of this methodology, using the Tons of Inductive
  Problems benchmark as a corpus.
\item Automated tooling to perform this benchmarking.
\item Application of our methodology to the QuickSpec and IsaCoSy MTE tools,
  and a comparison and discussion of the results.
\end{itemize}

@@

@@

The precision and recall of three MTE tools are compared
in~\cite{claessen2013automating}:
IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
IsaScheme~\cite{MontanoRivas2011} and
HipSpec~\cite{Claessen_hipspec:automating}.\footnote{Unfortunately this
  comparison only reports prior results for each tool, rather than reproducing
  them. This makes the numbers less comparable, since each tool was tested with
  slightly different definitions (e.g. whether or not the exponential and
  maximum functions were included).} The definitions and ground truths for all
three were taken from the standard library of the Isabelle proof assistant:
since the library authors have gone to the effort of stating and proving these
properties, and including them with every copy of their software, this is a good
indication that they are interesting. Two sets of definitions were used: a
theory of natural numbers; and a theory of lists (which we show in
Figure~\ref{figure:list_theory} on page~\pageref{figure:list_theory}). With only
two datapoints, each containing only a few definitions and properties (4
functions and 12 properties, in the natural number example), it is difficult
to draw general conclusions; yet despite these limitations we believe that this
is a viable method to performing meaningful, objective analyses and comparisons
for such tools. It is only with larger data sets that we can generalise the
measured performance to different, especially \emph{novel}, domains (which,
after all, is the purpose of MTE tools). In addition, we might be concerned that
such ``popular'' theories as the natural numbers may be unrepresentative of
typical usage. For example, we would not expect the library functions in a
typical verification problem to be as mathematically rich as number theory.

@@

The automation of mathematical tasks has been pursued since at least the time of
mechanical calculators like the Pascaline~\cite{d'ocagne}. A recurring theme in
these efforts is the separation between those undertaken by mathematicians like
Pascal and Babbage~\cite{bowden}, and those of engineers such as
M\"uller~\cite[p. 65]{lindgren}. This pattern continues today, with the tasks we
are concerned with (automatically constructing and evaluating concepts,
conjectures, theorems, axioms, examples, etc.) being divided into two main
fields: Mathematical Theory Exploration (MTE)~\cite{buchberger:06} (also
sometimes prefaced with ``Computer-Aided'', ``Automated'' or
``Algorithm-Supported''), which is championed by mathematicians such as
Buchberger~\cite{buchberger}; and Automated Theory Formation
(ATF)~\cite{lenat:77,colton:book}, pursued by AI researchers including Lenat.
Other related terms include ``Automated Mathematical
Discovery''~\cite{epstein:91,colton:interestingness,esarm2008},
``Concept Formation in Discovery Systems''~\cite{haase}, and
``Automated Theorem Discovery''~\cite{roy}.

@@

An important historical implementation of ATF is Lenat's AM (Automated
Mathematician) system. Unlike prior work, such as
Meta-Dendral~\cite{buchanan:75} and those described in~\cite{winston}, AM aims
to be a general purpose mathematical discovery system, designed to both
construct new concepts and conjecture relationships between them. AM is a
rule-based system which represents knowledge using a frame-like scheme, enlarges
its knowledge base via a collection of heuristic rules, and controls the firing
of these rules via an agenda mechanism. Evaluation of AM considered generality
(performance in new domains) and how finely-tuned various aspects of the program
are (the agenda, the interaction of the heuristics, etc). Most of this
evaluation was qualitative, and has subsequently been
criticised~\cite[chap.~13]{colton:book}. In their case study in methodology,
Ritchie and Hanna found a large discrepancy between the theoretical claims made
of AM and the implemented program~\cite{partridge}; for example, AM ``invented''
natural numbers from sets, but did so using a heuristic specifically designed to
make this connection.

@@

A similar ``layering'' approach is found in the IsaScheme system of
Monta{\~n}o-Rivas \etal{}~\cite{MontanoRivas2011}, which has also been
quantitatively compared against IsaCoSy and HipSpec using precision/recall
analysis~\cite{claessen2013automating}. The name comes from its embedding in the
Isabelle proof assistant and its use of ``schemes'': higher-order formulae which
can be used to generate new concepts and conjectures. Variables within a scheme
are instantiated automatically and this drives the invention process. For
example, the concept of ``repetition'' can be encoded as a scheme, and
instantiated with existing encodings of zero, successor and addition to produce
a definition of multiplication. The same scheme can be instantiated with this
new multiplication function to produce exponentiation.

>>>>

% Derived from the template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{csvsimple}
\usepackage{float}
\restylefloat{table}
\usepackage{hhline}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{arrows, calc, decorations.pathreplacing, positioning,
  shapes.geometric}
% etc.

@@

\begin{itemize}
\item A general methodology for benchmarking conjecture generation.
\item Resolving the issue of ``interestingness'' through the use of
  theorem-proving benchmarks as a ground-truth.
\item A specific instantiation of this methodology, using the Tons of Inductive
  Problems (TIP) benchmark as a corpus.
\item Automated tooling to perform this benchmarking.
\item Application of our methodology to the QuickSpec and IsaCoSy MTE tools,
  and a comparison and discussion of the results.
\end{itemize}

@@

\begin{sloppypar}
  Our benchmark, including the scripts used to generate it from the TIP problem
  set, is available online at
  \url{https://github.com/warbo/theory-exploration-benchmarks}. The application
  to QuickSpec is at \url{https://github.com/warbo/haskell-te} and IsaCoSy at
  \url{https://github.com/warbo/isaplanner-tip}.
\end{sloppypar}

@@

\begin{sloppypar}
  The precision and recall of three MTE tools are compared
  in~\cite{claessen2013automating}:
  IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
  IsaScheme~\cite{Montano-Rivas.McCasland.Dixon.ea:2012} and
  HipSpec~\cite{Claessen_hipspec:automating}.\footnote{Unfortunately this
    comparison only reports prior results for each tool, rather than reproducing
    them. This makes the numbers less comparable, since each tool was tested
    with slightly different definitions (e.g. whether or not the exponential and
    maximum functions were included).} The definitions and ground truths for all
  three were taken from the standard library of the Isabelle proof assistant:
  since the library authors have gone to the effort of stating and proving these
  properties, and including them with every copy of their software, this is a good
  indication that they are interesting. Two sets of definitions were used: a
  theory of natural numbers; and a theory of lists (which we show in
  Figure~\ref{figure:list_theory} on page~\pageref{figure:list_theory}). With
  only two datapoints, each containing only a few definitions and properties (4
  functions and 12 properties, in the natural number example), it is difficult
  to draw general conclusions; yet despite these limitations we believe that
  this is a viable method to performing meaningful, objective analyses and
  comparisons for such tools. It is only with larger data sets that we can
  generalise the measured performance to different, especially \emph{novel},
  domains (which, after all, is the purpose of MTE tools). In addition, we might
  be concerned that such ``popular'' theories as the natural numbers may be
  unrepresentative of typical usage. For example, we would not expect the
  library functions in a typical verification problem to be as mathematically
  rich as number theory.
\end{sloppypar}

@@

The automation of mathematical tasks has been pursued since at least the time of
mechanical calculators like the Pascaline~\cite{d'ocagne}. A recurring theme in
these efforts is the separation between those undertaken by mathematicians like
Pascal and Babbage~\cite{bowden}, and those of engineers such as
M\"uller~\cite[p. 65]{lindgren}. This pattern continues today, with the tasks we
are concerned with (automatically constructing and evaluating concepts,
conjectures, theorems, axioms, examples, etc.) being divided into two main
fields: Mathematical Theory Exploration (MTE)~\cite{buchberger:06} (also
sometimes prefaced with ``Computer-Aided'', ``Automated'' or
``Algorithm-Supported''), which is championed by mathematicians such as
Buchberger~\cite{buchberger}; and Automated Theory Formation
(ATF)~\cite{lenat:77,colton:book}, pursued by AI researchers including Lenat.
Other related terms include ``Automated Mathematical
Discovery''~\cite{epstein:91,colton2000notion,esarm2008},
``Concept Formation in Discovery Systems''~\cite{haase}, and
``Automated Theorem Discovery''~\cite{roy}.

@@

An important historical implementation of ATF is Lenat's AM (Automated
Mathematician) system. Unlike prior work, such as
Meta-Dendral~\cite{buchanan:75} and those described in~\cite{winston}, AM aims
to be a general purpose mathematical discovery system, designed to both
construct new concepts and conjecture relationships between them. AM is a
rule-based system which represents knowledge using a frame-like scheme, enlarges
its knowledge base via a collection of heuristic rules, and controls the firing
of these rules via an agenda mechanism. Evaluation of AM considered generality
(performance in new domains) and how finely-tuned various aspects of the program
are (the agenda, the interaction of the heuristics, etc). Most of this
evaluation was qualitative, and has subsequently been
criticised~\cite[chap.~13]{colton:book}. In their case study in methodology,
Ritchie and Hanna found a large discrepancy between the theoretical claims made
of AM and the implemented program~\cite{ritchie1984case}; for example, AM ``invented''
natural numbers from sets, but did so using a heuristic specifically designed to
make this connection.

@@

A similar ``layering'' approach is found in the IsaScheme system of
Monta{\~n}o-Rivas \etal{}~\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, which
has also been quantitatively compared against IsaCoSy and HipSpec using
precision/recall analysis~\cite{claessen2013automating}. The name comes from its
embedding in the Isabelle proof assistant and its use of ``schemes'':
higher-order formulae which can be used to generate new concepts and
conjectures. Variables within a scheme are instantiated automatically and this
drives the invention process. For example, the concept of ``repetition'' can be
encoded as a scheme, and instantiated with existing encodings of zero, successor
and addition to produce a definition of multiplication. The same scheme can be
instantiated with this new multiplication function to produce exponentiation.
